Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-15 Paper_Draft2/Draft2.Rmd

author:
  - name: Amrita Acharya
    affil: 1, 2
  - name: Dianne Caravela
    affil: 1, 3
  - name: Eunice Kim
    affil: 1, 4
  - name: Emma Kornberg
    affil: 1, 5
  - name: Elisabeth Nesmith
    affil: 1, 6
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
  - num: 2
    email: aacharya@smith.edu
  - num: 3
    email: dcaravela@smith.edu
  - num: 4
    email: ekim89@smith.edu
  - num: 5
    email: ekornberg@smith.edu
  - num: 6
    email: enesmith@smith.edu
journal: water
type: article
status: submit
bibliography: mybibfile.bib
csl: ../resources/chicago-author-date.csl
#appendix: appendix.tex
classoption: fleqn

@@

@@

@@

**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between .8 to 1.25 is considered fair:

@@

**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants.

@@

**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group.

The ideal value is 0. A value less than 0 implies higher benefit for the privileged group and a value greater than 0 implies higher benefit for the unprivileged group.
Fairness for this metric is between -0.1 and 0.1.
This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic. This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.

@@

For the next step of our experiment, we needed to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms. These are all based on using predictive models to figure out how we can “fix” the bias that is present. 
Pre-processing refers to assessing the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data. It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection. With pre-processing, we use the re-weighing algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018]. We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context. After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.

Another approach, in-processing, explores how bias can be targeted while building a model. It assesses how we build learning algorithms, which often prioritize accuracy over fairness. The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric. In order to do this, we first needed to convert our data frame into a data type called a . Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model outputs. Similarly to the pre-processing, it is flexible because it does not work directly with the algorithm, but instead has direct access to all the predicted classification values. Because the algorithms are often so complex and difficult to interpret, post-processors are trained in a black box. To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set. Based on these results, we looked comparatively at all the approaches to determine which we assessed to be the most fair.

@@

# Discussion

# Conclusion

>>>>

author:
  - name: Amrita Acharya
    affiliation: 1, 2
  - name: Dianne Caravela
    affiliation: 1, 3
  - name: Eunice Kim
    affiliation: 1, 4
  - name: Emma Kornberg
    affiliation: 1, 5
  - name: Elisabeth Nesmith
    affiliation: 1, 6
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
  - num: 2
    email: aacharya@smith.edu
  - num: 3
    email: dcaravela@smith.edu
  - num: 4
    email: ekim89@smith.edu
  - num: 5
    email: ekornberg@smith.edu
  - num: 6
    email: enesmith@smith.edu
journal: water
type: article
status: submit
bibliography: mybibfile.bib
csl: ../resources/chicago-author-date.csl
#appendix: appendix.tex
classoption: fleqn

@@

```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'The race distribution for black people is 3000 and the distribution for white people is 2100. Therefore, there are more black people in the dataset than white people.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_bar_plot.png")
```

```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot is a Kernel Density Estimate (KDE) plot. The y-axis represents the KDE of the Probability Distribution Function of the age variable in our dataset. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line shows the average age of Black defendants and the green dotted line shows the average age of white defendants. For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_age_plot.png")
```

```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'Of both black and white people, there is a higher percentage of people who did not recidivate than who did. 2500 people recidivated, whereas approximately 2600 did not.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/recid_bar_plot.png")
```

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants. Black defendants recidivate are more likely than not to recidivate. However, we can also see that there are more Black defendants in the dataset in general.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_recid_plot.png")

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to "to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms" [@aif360-oct-2018].

helps facilitate the examination of algorithmic fairness. It contains multiple data sets, including the COMPAS data set that accompanied the Angwin et al. piece, and provides

@@

**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other. In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate. A value of 1 means that the ratio is exactly 1:1. Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits. According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].

@@

**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].

@@

**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].

The ideal value is 0. A value less than 0 implies higher benefit for the privileged group and a value greater than 0 implies higher benefit for the unprivileged group.
Fairness for this metric is between -0.1 and 0.1 [@aif360-oct-2018].

This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How]. This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.

@@

For the next step of our experiment, we need to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms. These are all based on using predictive models to figure out how we can “fix” the bias that is present.

Pre-processing refers to assessing the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data. It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection. With pre-processing, we use the re-weighing algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018]. We use a logistic regression model for this algorithm, as it is the easiest to interpret in the given context. After running the fairness metrics using the pre-processing algorithm, we are able to compare our results to the baseline metrics from the previous section.

Another approach, in-processing, explores how bias can be targeted while building a model. It assesses how we build learning algorithms, which often prioritize accuracy over fairness. The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric. Similarly to pre-processing, we compare the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model outputs. Similarly to the pre-processing, it is flexible because it does not work directly with the algorithm, but instead has direct access to all the predicted classification values. Because the algorithms are often so complex and difficult to interpret, post-processors are trained in a black box. To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set [@aif360-oct-2018]. Based on these results, we look comparatively at all the approaches to determine which we assess to be the most fair.

@@

                                 Result   Benefit   
                                +---------------+
Statistical Parity Difference   | -0.14 | white |
Disparate Impact Ratio          | 0.47  | white |
Average Odds Difference         | -0.44 | white |
Equal Opportunity Difference    | -0.41 | white |
                                +---------------+
